---
title: "Module 2.1, Lecture 2.1 Notes"
author: "Jelena H. Pantel"
institute: "University of Duisburg-Essen"
date: "`r file.mtime(knitr::current_input())`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(deSolve)
```

### Deriving the recursion equations (Otto & Day Ch 2, equation 2.4)
Using the life cycle diagram (Figure 2.2b), we begin from:

$$\begin{eqnarray} n'(t) = n(t) - dn(t) \\ n''(t) = n'(t) + bn'(t) \\ n'''(t) = n''(t) + m  \end{eqnarray}$$
1. Substitute $n'(t)$ equation into $n''(t)$ equation:

$$ n''(t) = n(t) - dn(t) + b(n(t) - dn(t)) $$

Using the logic of $1(x) + 2(x) = (1+2)x$, we re-arrange the right side to:

$$ 1(n(t) - dn(t)) + b(n(t)-dn(t)) = (1+b)(n(t) - dn(t)) $$
The $n(t)-dn(t)$ part can be re-written as $(1-d)n(t)$, leaving:

$$(1+b)(1+d)n(t)$$

2. Substitute $n''i(t)$ equation into $n'''(t)$ equation:

$$ n'''(t) = n(t+1) = (1+b)(1-d)n(t)-m $$
## Some questions to answer before class

**1. Explain the following statement- *R*, the number of surviving individuals per parent, is constant regardless of population size, and invariable for all individuals in the population. Contrast this with *age-structured* models (like we discussed last week).**

We recall from Jackson et al. 2000 that in an age-structured model, not all individuals participate equally in the processes. In that example, they didn’t all have identical birth, predation, mortality rates.

**2. Explain the difference between *R* in the recursion equation for exponential growth and *r* in the differential equation. What does *per capita* mean? What value of *R* indicates the population size is not changing and what value of *r* indicates the population size is not changing?**

*R* in the recursion equation (discrete-time) is – the number of individuals that each parent / *N* is replaced by in the next time unit. A population would be considered to be constant (not changing) if each parent / *N* replaces itself… and thus *R* = 1.

We can see this in R: see R code, Exercise 7

```{r}
## Exercise 7. Simulation model of discrete-time exponential growth
R <- 1
N0 <- 4
disc_exp <- function(R,N0){
  Nt1 <- R*N0
  return(Nt1)
}
N <- N0

for(i in 1:100){
  Nt1 <- disc_exp(R,N0)
  N <- c(N,Nt1)
  N0 <- Nt1
}
plot(N,ylab="time")
```

In the differential / continuous time model (if births and deaths can occur at any time point, rather than in specific seasons), the population growth rate term is instead *r*, which is *r*=*b*-*d*, or the per capita rate of change in the number of individuals. This tells us the rate at which the population changes per individual in the population - the *r* value can be multiplied by each individual in the population to determine their total contribution to population growth. For example, if *r* = 0, then births are equal to deaths, and so the population is not changing. We refer to *r* as the per capita change in the number of individuals because the value is associated with each individual in the population.

We can see this in R: R code, Exercise 8

```{r}
## Exercise 8. Continuous time model of exponential growth
library(deSolve)
cont_exp <- function(t, state, parameters) {
  with(as.list(c(state, parameters)), {
    dN <-  r * N
    return(list(dN))
  })
}

parameters <- c(r = 0)
state      <- c(N = 42)
times      <- seq(0, 100, by = 0.01)

out <- deSolve::ode(y = state, times = times, func = cont_exp, parms = parameters)

plot(out)
```

**3. How did the authors come up with the equation above Figure 3.3 directly from that figure? (Recall the formula for a line)** $y=mx+b$

The logistic growth model arises when we consider that the population growth rate *r* is not fixed, but instead is density-dependent. It decreases as the population size approaches carrying capacity *K*. So we see a graph where *r* is a linear function of *N*. Since the relationship between *r* and *N* is linear, we can use the formula of a line to determine this function. Recall $y = mx + b$, where *m* is the slope and *b* is the y-intercept.

We see from the plot two values of *R*: *R* = 1 when *N* = *K* (why? because population size doesn’t change anymore when $N = K$), and $R = 1+r_d$ when *N* = 0 (because the population is at its maximal growth rate when there is no intra-specific competition). This gives us the y-intercept. To get the slope, we look at the change in *y* over the change in *x*. We take 2 points, $(x_1,y_1)$  and $(x_2,y_2)$, which can be $(0,1+r_d)$ and $(K,1)$. To get the slope, its $\frac {y_2-y_1} {x_2-x_1}$, and that is: $\frac {1 – (1 + rd)} {K – 0} = \frac {-r_d} {K}$. We put that together to get: $R(n) = (1 + r_d) + \frac {-r_d}{K}n(t)$.

By the way, if you want to go on to derive the discrete-time logistic growth equation:

Begin from the logistic growth equation:

$$ n(t+1) = Rn(t) $$

And substitute the *new* value of R we derived above:

$$ n(t+1) = [(1+r_d) + \frac{-r_d}{K}n(t)]n(t) = n(t) + r_dn(t) + r_dn(t)(\frac{-n(t)}{K}) = n(t) + r_dn(t)(1-\frac{n(t)}{K})$$

**4. Be prepared to repeat what's written in the Box under equation 3.5b when I ask "what does the *logistic growth model assume about how population growth rate changes with increasing population size?"**

The logistic model assumes that the rate of change of a variable decreases linearly as the variable increases in value, with no change occurring if the variable is at the carrying capacity *K*.

**5. Explain the difference between a linear and non-linear model of population growth (exponential and logistic), basically using what is written in the box under Figure 3.4, and using equations 3.5b and 3.2 to illustrate this difference.**

In a linear model, the change per unit time is a *linear* function of the variables; models that involve more complicated functions of the variables are called non-linear models.

We used $(R-1) = r_d$, so this is the per capita change in the number of individuals from one generation to the next. Equation 3.2 shows us that changes in *n* are a linear function of $r_d$ and $n(t)$. In equation 3.5b, change in *n* depends on $n(t)$ in a non-linear way.

Begin from: $R = 1 + r_d$

equation 3.2) $\Delta n = (R-1)n(t)$ --> linear


equation 3.5b) $\Delta n = r_dn(t)(1 - \frac{n(t)}{K})$ --> non-linear

Where $\Delta n$ --> a difference equation, $\Delta n = n(t+1) - n(t)$

We can see this in R: R code, Exercise 9

```{r}
## Exercise 9. Linear and non-linear models
models <- function(rd,nt,K){
  dN.1 <- rd * nt
  dN.2 <- rd * nt * (1 - nt/K)
  return(list(dN.1,dN.2))
}

rd <- 1.7
nt <- seq(100,700000,by=1000)
K <- 75000

res <- models(rd,nt,K)
plot(nt,res[[1]],ylim=c(-10000000,10000000),xlab="n(t)",ylab="delta-N")
points(nt,res[[2]])
```



