---
title: "Week 8 Reading Guidelines"
author: "Jelena H. Pantel"
institute: "University of Duisburg-Essen"
date: "`r file.mtime(knitr::current_input())`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Notes
This is the reading guide for Jager et al. 2011 and Le et al. 2011.

## Some questions to answer before class

1. Toxicology labs exist that can measure the effects of substances on different organisms - what then is the advantage gained by considering an ecotoxicology model? (hint: see Introduction of Jager et al. 2011)

2. Please define *toxokinetics* (hint: see Introduction of Jager et al. 2011)

3. When plotting data in general, we must isolate the *causal variable* and the *response variable*. For example, I might be interested in the effects of “Sleeping past my alarm” on “Missing the bus”. I have a causal variable *x*, defined as the number of minutes past 7am that I sleep in. I have a response variable *y*, defined as the probability that I miss my bus. I don’t know the relationship between these two perfectly, but I do know (1) that the relationship is positive (the longer I sleep in, the more likely I am to miss my bus) and (2) that the relationship is non-linear (if I sleep in for 5-10 minutes, I usually catch my bus, but if I sleep in from 20-30 minutes I often miss the bus. If I sleep in past 40 minutes, I always miss my bus.

```{r echo=FALSE}
x <- seq(0,60)
S <- 1 / (1 + exp(-.4*(x-30)))
plot(x,S,type="l",lwd=3,xlab="Minutes slept in", ylab="probability of missing bus")
```

Now to the question! In Figure 1 of Jager et al (the top part of the figure, with boxes that include toxicokinetics and toxicodynamics), you see a causal variable on the far left and a response variable on the far right. Would you expect the relationship between these to be negative or positive? Linear, or non-linear?

For a non-linear relationship, please go through Jager et al. and come up with at least one reason why the response should be non-linear.

4. What data is needed to estimate the parameters in a TKTD model (see Jager et al discussion)?

5. What do you prefer: one- or multiple-compartment models to describe toxicokinetics?

6. In your opinion, which processes play an important role in toxicokinetics of metals, but not that of organic compound?

7. Similarly, which processes are important to toxicokinetics of organic chemicals, but not to that of metals?

8. How do you define a model as mechanistic?

9. In the paper from Jager et al., the authors mentioned two assumptions: individual tolerance and stochastic death. Based on the explanation by the authors, how would you interpret an assumption that combines both of these terms?
